Perturbing SU($n$)-symmetric interactions

### 2019-10-30 collective_sun/sun_perturbation.tex
adding table of contents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\thispagestyle{fancy}

@@

We consider an array of $N$ multilevel spins with non-uniform
SU($n$)-symmetric interactions that can be written in the form
\begin{align}
  H_0 = \sum_{p<q} h_{pq} \Pi_{pq},
  &&
  \Pi_{pq} \equiv \sum_{\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)},
\end{align}
where $S_{\mu\nu}^{(p)}\equiv\op{\mu}{\nu}_p$ flips the state of spin
$p$ to $\ket\mu$ from $\ket\nu$; the operator $\Pi_{pq}$ permutes
spins $p$ and $q$, and the coefficients $h_{pq}$ satisfy
\begin{align}
  h_{pp} = 0
  &&
  h_{pq} \le 0,
  &&
  h_{pq} = h_{qp}.
\end{align}
The ground-state manifold of the interaction Hamiltonian $H_0$
consists of fully symmetric states that are simultaneous $+1$
eigenstates of all permutation operators $\Pi_{pq}$.  The energy of
these states is
\begin{align}
  E_0 \equiv \sum_{p<q} h_{pq} = \f12 \sum_{p,q} h_{pq}
  = \f12 \sum_p h_p = \f12 N h,
\end{align}
where we define
\begin{align}
  h_p \equiv \sum_q h_{pq},
  &&
  h \equiv \EE_p\sp{h_p} = \f1N \sum_{p,q} h_{pq},
  &&
  \EE_p\sp{X_p} \equiv \f1N \sum_p X_p.
\end{align}
We wish to determine the effective dynamics induced on the manifold
$\M_0$ of fully symmetric states by weak perturbations of the form
\begin{align}
  \V_1 \equiv \sum_p v_p V_p,
  &&
  \V_2 \equiv \sum_{p<q} v_{pq} V_p V_q,
  \label{eq:perturbations}
\end{align}
where $V_p$ is a single-body operator on spin $p$, and
$v_{pq}=v_{qp}$.  The effective Hamiltonian $H_W$ induced on the
ground-state manifold $\M_0$ by a perturbation $\V_W$ through second
order in perturbation theory is given by\cite{bravyi2011schrieffer,
  perlin2019effective}
\begin{align}
  H_W = H_W^{(1)} + H_W^{(2)},
  &&
  H_W^{(1)} = \P_0 \V_W \P_0,
  &&
  H_W^{(2)} = - \P_0 \V_W \E \V_W \P_0,
  &&
  \E \equiv \sum_{k>0} \f{\P_k}{\Delta_k},
\end{align}
where $\P_k$ is a projector onto the $k$-th excited eigenspace of the
interaction Hamiltonian $H_0$, with interaction energy $\Delta_k$
above that of fully symmetric manifold $\M_0$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Finding eigenstates of the interaction Hamiltonian}

@@

The two-body case requires a bit of work to re-cast as an
eigenvector/eigenvalue problem.  We first write the matrix $\m v$ as a
vector,
\begin{align}
  \m v = \sum_{p,q} v_{pq} \op{k}{q}
  \to \v{\m v} \equiv \sum_{p,q} v_{pq} \ket{pq},
\end{align}
and define the matrix
\begin{align}
  \check{\m h}
  \equiv \sum_{k,\ell} \bar h_{k\ell} \op{k\ell}
  + \sum_{k,q,\ell} h_{kq} \op{k\ell}{q\ell},
  \label{eq:h_super_mat}
\end{align}
in terms of which the condition in \eqref{eq:v_cond_2} becomes
\begin{align}
  \check{\m h} \c \v{\m v} = \e\p{\m v} \v{\m v}.
\end{align}
Note that the matrix $\check{\m h}$ in \eqref{eq:h_super_mat} can be
written in the block-diagonal form
\begin{align}
  \check{\m h} = \sum_\ell \m h_\ell \otimes \op{\ell},
  &&
  \m h_\ell
  \equiv \sum_{k,q}\p{\delta_{kq} \bar h_{k\ell} + h_{kq}} \op{k}{q},
\end{align}
where each block $\m h_\ell$ is an $N\times N$ matrix.  We can
diagonalize these blocks independently, and classify their
eigenvectors by corresponding eigenvalues.  ``Total'' eigenvectors
$\v{\m v}$ of $\check{\m h}$ are then essentially concatenations of
eigenvectors from within each block $\m h_\ell$ that all share the
same eigenvalue.  The task of constructing eigenvectors
$\V_2\p{\m v}\ket\psi$ of $H_0$ and determining their energy thus
reduces to finding the eigenvectors and eigenvalues of $N$ matrices
$\m h_\ell$ of size $N\times N$.  If the interaction Hamiltonian $H_0$
is translationally invariant, then $\bar h_{k\ell}=h_{k\ell}-h$, and
$h_{k\ell}$ depends only on the displacement between the spins indexed
by $k$ and $\ell$.  The blocks $\m h_\ell$ then take the form
\begin{align}
  \m h_{\ell} = \m\omega_\ell \c \m h_0 \c \m\omega_\ell^\dag,
  &&
  \m\omega_\ell \equiv \sum_k \op{k+\ell}{k},
  \label{eq:block_shift}
\end{align}
which implies that we need only find the eigenvectors and eigenvalues
of a single block $\m h_0$, after which we can determine the
eigenvectors of any block simply by appropriate applications of the
shift matrix $\m\omega_\ell$.  Note that in dimensions $D>1$, the
relation in \eqref{eq:block_shift} holds for vectors $k,\ell$ that
index lattice sites, i.e.~$k=\p{k_1,k_2,\cdots,k_D}$ and likewise with
$\ell$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

\tableofcontents

\section{Introduction}

We consider an array of $N$ multilevel spins with non-uniform
SU($n$)-symmetric interactions that can be written in the form
\begin{align}
  H_0 = \sum_{p<q} h_{pq} \Pi_{pq},
  &&
  \Pi_{pq} \equiv \sum_{\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)},
\end{align}
where $S_{\mu\nu}^{(p)}\equiv\op{\mu}{\nu}_p$ flips the state of spin
$p$ to $\ket\mu$ from $\ket\nu$; the operator $\Pi_{pq}$ permutes
spins $p$ and $q$; and the coefficients $h_{pq}$ satisfy
\begin{align}
  h_{pp} = 0
  &&
  h_{pq} \le 0,
  &&
  h_{pq} = h_{qp}.
\end{align}
The ground-state manifold of the interaction Hamiltonian $H_0$
consists of fully symmetric states that are simultaneous $+1$
eigenstates of all permutation operators $\Pi_{pq}$.  The energy of
these states is
\begin{align}
  E_0 \equiv \sum_{p<q} h_{pq} = \f12 \sum_{p,q} h_{pq}
  = \f12 \sum_p h_p = \f12 N h,
\end{align}
where we define
\begin{align}
  h_p \equiv \sum_q h_{pq},
  &&
  h \equiv \EE_p\sp{h_p} = \f1N \sum_{p,q} h_{pq},
  &&
  \EE_p\sp{X_p} \equiv \f1N \sum_p X_p.
\end{align}
We wish to determine the effective dynamics induced on the manifold
$\M_0$ of fully symmetric states by weak perturbations of the form
\begin{align}
  \V_1 \equiv \sum_p v_p V_p,
  &&
  \V_2 \equiv \sum_{p<q} v_{pq} V_p V_q,
  \label{eq:perturbations}
\end{align}
where $V_p$ is a single-body operator on spin $p$, and
$v_{pq}=v_{qp}$.  The effective Hamiltonian $H_W$ induced on the
ground-state manifold $\M_0$ by a perturbation $\V_W$ through second
order in perturbation theory is given by\cite{bravyi2011schrieffer,
  perlin2019effective}
\begin{align}
  H_W = H_W^{(1)} + H_W^{(2)},
  &&
  H_W^{(1)} = \P_0 \V_W \P_0,
  &&
  H_W^{(2)} = - \P_0 \V_W \E \V_W \P_0,
  &&
  \E \equiv \sum_{k>0} \f{\P_k}{\Delta_k},
\end{align}
where $\P_k$ is a projector onto the $k$-th excited eigenspace of the
interaction Hamiltonian $H_0$, with interaction energy $\Delta_k$
above that of fully symmetric manifold $\M_0$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Constructing eigenstates of SU($n$)-symmetric interactions}

@@

The two-body case requires a bit of work to re-cast as an
eigenvector/eigenvalue problem.  We first write the matrix $\m v$ as a
vector,
\begin{align}
  \m v = \sum_{k,\ell} v_{k\ell} \op{k}{\ell}
  \to \v{\m v} \equiv \sum_{k,\ell} v_{k\ell} \ket{k\ell},
\end{align}
and define the matrix
\begin{align}
  \check{\m h}
  \equiv \sum_{k,\ell} \bar h_{k\ell} \op{k\ell}
  + \sum_{k,q,\ell} h_{kq} \op{k\ell}{q\ell},
  \label{eq:h_super_mat}
\end{align}
in terms of which the condition in \eqref{eq:v_cond_2} becomes
\begin{align}
  \check{\m h} \c \v{\m v} = \e\p{\m v} \v{\m v}.
\end{align}
Note that the matrix $\check{\m h}$ in \eqref{eq:h_super_mat} can be
written in the block-diagonal form
\begin{align}
  \check{\m h} = \sum_\ell \m h_\ell \otimes \op{\ell},
  &&
  \m h_\ell
  \equiv \sum_{k,q}\p{\delta_{kq} \bar h_{k\ell} + h_{kq}} \op{k}{q},
\end{align}
where each block $\m h_\ell$ is an $N\times N$ matrix.  We can
diagonalize these blocks independently, and classify their
eigenvectors by the corresponding eigenvalues.  ``Total'' eigenvectors
$\v{\m v}$ of $\check{\m h}$ are then essentially built from
eigenvectors within each block $\m h_\ell$ that share the same
eigenvalue.  The task of constructing eigenvectors
$\V_2\p{\m v}\ket\psi$ of $H_0$ and determining their energy thus
reduces to finding the eigenvectors and eigenvalues of $N$ matrices
$\m h_\ell$ of size $N\times N$.

If the interaction Hamiltonian $H_0$ is translationally invariant,
then $\bar h_{k\ell}=h_{k\ell}-h$, and $h_{k\ell}$ depends only on the
displacement between the spins indexed by $k$ and $\ell$.  The blocks
$\m h_\ell$ then take the form
\begin{align}
  \m h_{\ell} = \m\omega_\ell \c \m h_0 \c \m\omega_\ell^\dag,
  &&
  \m\omega_\ell \equiv \sum_k \op{k+\ell}{k},
  \label{eq:block_shift}
\end{align}
which implies that we need only find the eigenvectors and eigenvalues
of a single block $\m h_0$, after which we can determine the
eigenvectors of any block simply by appropriate applications of the
shift matrix $\m\omega_\ell$.  Note that in dimensions $D>1$, the
relation in \eqref{eq:block_shift} holds for vectors $k,\ell$ that
index lattice sites, i.e.~$k=\p{k_1,k_2,\cdots,k_D}$ and likewise with
$\ell$.
